---
title: "MMM_Coffee_shop"
author: "Elena Marochkina"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

``` {r}
# Add Libraries 
library(readxl)
library(skimr)
library(tidyverse)
library(tidyr)
library(dplyr)
library(flextable)
library(officer)
library(lubridate)
library(ggplot2)

# Set the custom theme
theme_custom <- theme(
    plot.title = element_text(size = 15, hjust = 0.5),
    plot.subtitle = element_text(size = 15, hjust = 0.5),
    strip.text = element_text(size = 10),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 15),
    legend.title = element_text(size = 15),
    legend.text = element_text(size = 10),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    
  )
```

# Read Data and Set Variables Types

```{r coffee shop data}
# Read data and transform categorical variables to factors
data_sales <- read_excel("data/raw/Coffee_Shop_Sales.xlsx") %>%
  mutate(across(where(is.character), as.factor),
         across(c(transaction_id, store_id, product_id), as.factor))

# Take a look on data basic statistics
skim(data_sales)
```

# Clean Data and Check for Outliers

```{r clean data, width = 24, height = 6}
# Check for missing data
missing_data <- data_sales %>%
  summarise(across(everything(), ~ sum(is.na(.))))  # Count missing values in each column

# Check for duplicates
duplicates <- data_sales %>%
  filter(duplicated(.))

# Detect outliers in unit_price within each product_id group
detect_outliers_by_product <- function(column) {
  mean_val <- mean(column, na.rm = TRUE)
  sd_val <- sd(column, na.rm = TRUE)
  
  # Define bounds using the three-sigma rule
  lower_bound <- mean_val - 3 * sd_val
  upper_bound <- mean_val + 3 * sd_val
  
  # Return outliers
  outliers <- column < lower_bound | column > upper_bound
  return(outliers)
}

# Apply outlier detection for unit_price by product_id
outliers_by_product_price <- data_sales %>%
  group_by(product_id) %>%
  mutate(price_outliers = detect_outliers_by_product(unit_price)) %>%

  # Filter for products that have price outliers
  filter(any(price_outliers)) %>%
  ungroup()

# Visualize the outliers
# Boxplot for unit_price by product_id
ggplot(outliers_by_product_price, aes(x = as.factor(product_id), y = unit_price)) +
  geom_boxplot(width = 0.9) +
  labs(title = "Boxplot of Unit Price by Product ID", x = "Product ID", y = "Unit Price") +
  coord_flip()+
  theme_custom +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

Upon analysis, it was determined that outlier detection for *transaction_qty* is unnecessary, as the observed range (1–8) reflects typical and reasonable purchase quantities.

Similarly, no outliers were removed for *unit_price*, as all values, including price reductions of up to 50% (price_id = 9), are consistent with pricing strategies such as discounts or promotions.

Given the acceptability of the data, no further outlier treatment is required for either *transaction_qty* or *unit_price*, ensuring that the dataset remains representative of normal business operations.

# Desriptive Statistics
## Descriptive Statistics for Numerical Variables

```{r descriptive statistics for numerical vars}
# Function to calculate 95% confidence interval for the mean
ci_95 <- function(x) {
  n <- sum(!is.na(x))
  if (n < 3) return("NA")
  se <- sd(x, na.rm = TRUE) / sqrt(n)
  mean_x <- mean(x, na.rm = TRUE)
  ci <- c(mean_x - 1.96 * se, mean_x + 1.96 * se)
  paste0(round(ci[1], 2), " - ", round(ci[2], 2))
}

# List of descriptive statistics
statistics <- list(
  `_Number of values` = ~as.character(sum(!is.na(.x))),
  `_No data` = ~as.character(sum(is.na(.x))),
  `_Mean` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(mean(.x, na.rm = TRUE) %>% round(2))),
  `_Median` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(median(.x, na.rm = TRUE) %>% round(2))),
  `_SD` = ~ifelse(sum(!is.na(.x)) < 3, "NA", as.character(sd(.x, na.rm = TRUE) %>% round(2))),
  `_Q1 - Q3` = ~ifelse(sum(!is.na(.x)) == 0, "NA", paste0(as.character(quantile(.x, 0.25, na.rm = TRUE) %>% round(2)), " - ", as.character(quantile(.x, 0.75, na.rm = TRUE) %>% round(2)))),
  `_IQR` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(IQR(.x, na.rm = TRUE) %>% round(2))),
  `_95% CI` = ~ci_95(.x),
  `_min` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(min(.x, na.rm = TRUE) %>% round(2))),
  `_max` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(max(.x, na.rm = TRUE) %>% round(2)))
)

# Summarize the statistics for each numeric variable grouped by TenYearCHD
data_sales %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), statistics)) %>%
  pivot_longer(cols = everything(), names_sep = "__", names_to = c("Variable", "Stat")) %>%
  rename(`Value` = value) %>%
  flextable() %>%
  theme_zebra() %>%
  merge_v("Variable") %>%
  width(j = c("Variable", "Stat", "Value"), width = 2)
```

## Descriptive Statistics for Categorical Variables

```{r}
# Clean and summarize the categorical data for all factor variables
data_sales %>%
  select(-transaction_id, -product_id, -product_detail) %>%
  select(where(is.factor)) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable, Value) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(Variable) %>%
  mutate(`No data` = sum(is.na(Value)),  # Calculate the number of missing values for each variable
         `% by group` = (n / sum(n)) * 100) %>%
  ungroup() %>%
  select(Variable, Value, n, `% by group`, `No data`) %>%
  arrange(Variable, Value) %>%
  flextable() %>%
  theme_zebra() %>%
  merge_v("Variable")
```

# Analyze Data
## Sales By Product Category
```{r}

```
