---
title: "MMM_Coffee_shop"
author: "Elena Marochkina"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Add Libraries 
library(readxl)
library(skimr)
library(tidyverse)
library(tidyr)
library(dplyr)
library(flextable)
library(lubridate)
library(ggplot2)
library(gridExtra)
library(scales)

# Set the custom theme
theme_custom <- theme_minimal() +
  theme (
    plot.title = element_text(size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    strip.text = element_text(size = 10),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    
  )
```

# 1.    Read Data and Set Variables Types

```{r coffee shop data}
# Read data and transform categorical variables to factors
data_sales <- read_excel("data/raw/Coffee_Shop_Sales.xlsx") %>%
  mutate(across(where(is.character), as.factor),
         across(c(transaction_id, store_id, product_id), as.factor))

# Take a look on data basic statistics
skim(data_sales)
```

# 2.    Clean Data and Check for Outliers

```{r clean data, width = 24, height = 6}
# Check for missing data
missing_data <- data_sales %>%
  summarise(across(everything(), ~ sum(is.na(.))))  # Count missing values in each column

# Check for duplicates
duplicates <- data_sales %>%
  filter(duplicated(.))

# Detect outliers in unit_price within each product_id group
detect_outliers_by_product <- function(column) {
  mean_val <- mean(column, na.rm = TRUE)
  sd_val <- sd(column, na.rm = TRUE)
  
  # Define bounds using the three-sigma rule
  lower_bound <- mean_val - 3 * sd_val
  upper_bound <- mean_val + 3 * sd_val
  
  # Return outliers
  outliers <- column < lower_bound | column > upper_bound
  return(outliers)
}

# Apply outlier detection for unit_price by product_id
outliers_by_product_price <- data_sales %>%
  group_by(product_id) %>%
  mutate(price_outliers = detect_outliers_by_product(unit_price)) %>%

  # Filter for products that have price outliers
  filter(any(price_outliers)) %>%
  ungroup()

# Visualize the outliers
# Boxplot for unit_price by product_id
ggplot(outliers_by_product_price, aes(x = as.factor(product_id), y = unit_price)) +
  geom_boxplot(width = 0.9) +
  labs(title = "Boxplot of Unit Price by Product ID", x = "Product ID", y = "Unit Price") +
  coord_flip()+
  theme_custom 

```

Upon analysis, it was determined that outlier detection for *transaction_qty* is unnecessary, as the observed range (1–8) reflects typical and reasonable purchase quantities.

Similarly, no outliers were removed for *unit_price*, as all values, including price reductions of up to 50% (price_id = 9), are consistent with pricing strategies such as discounts or promotions.

Given the acceptability of the data, no further outlier treatment is required for either *transaction_qty* or *unit_price*, ensuring that the dataset remains representative of normal business operations.

# 3.    Desriptive Statistics

## 3.1.   Descriptive Statistics for Numerical Variables

```{r numerical variables}
# Function to calculate 95% confidence interval for the mean
ci_95 <- function(x) {
  n <- sum(!is.na(x))
  if (n < 3) return("NA")
  se <- sd(x, na.rm = TRUE) / sqrt(n)
  mean_x <- mean(x, na.rm = TRUE)
  ci <- c(mean_x - 1.96 * se, mean_x + 1.96 * se)
  paste0(round(ci[1], 2), " - ", round(ci[2], 2))
}

# List of descriptive statistics
statistics <- list(
  `_Number of values` = ~as.character(sum(!is.na(.x))),
  `_No data` = ~as.character(sum(is.na(.x))),
  `_Mean` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(mean(.x, na.rm = TRUE) %>% round(2))),
  `_Median` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(median(.x, na.rm = TRUE) %>% round(2))),
  `_SD` = ~ifelse(sum(!is.na(.x)) < 3, "NA", as.character(sd(.x, na.rm = TRUE) %>% round(2))),
  `_Q1 - Q3` = ~ifelse(sum(!is.na(.x)) == 0, "NA", paste0(as.character(quantile(.x, 0.25, na.rm = TRUE) %>% round(2)), " - ", as.character(quantile(.x, 0.75, na.rm = TRUE) %>% round(2)))),
  `_IQR` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(IQR(.x, na.rm = TRUE) %>% round(2))),
  `_95% CI` = ~ci_95(.x),
  `_min` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(min(.x, na.rm = TRUE) %>% round(2))),
  `_max` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(max(.x, na.rm = TRUE) %>% round(2)))
)

# Summarize the statistics for each numeric variable grouped by TenYearCHD
data_sales %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), statistics)) %>%
  pivot_longer(cols = everything(), names_sep = "__", names_to = c("Variable", "Stat")) %>%
  rename(`Value` = value) %>%
  flextable() %>%
  theme_zebra() %>%
  merge_v("Variable") %>%
  width(j = c("Variable", "Stat", "Value"), width = 2)
```

## 3.2.   Descriptive Statistics for Categorical Variables

```{r categorical variables}
# Clean and summarize the categorical data for all factor variables
data_sales %>%
  select(-transaction_id, -product_id, -product_detail) %>%
  select(where(is.factor)) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable, Value) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(Variable) %>%
  mutate(`No data` = sum(is.na(Value)),  # Calculate the number of missing values for each variable
         `% by group` = (n / sum(n)) * 100) %>%
  ungroup() %>%
  select(Variable, Value, n, `% by group`, `No data`) %>%
  arrange(Variable, Value) %>%
  flextable() %>%
  theme_zebra() %>%
  merge_v("Variable")
```

# 4.    Data Analysis
## 4.1.   Sales By Product
### 4.1.1.    Sales By Product Category

```{r product category}
# Aggregate the total sales by product category
sales_by_category <- data_sales %>%
  select(product_category, transaction_qty, unit_price) %>%
  group_by(product_category) %>%
  summarise(total_sales_qty = sum(transaction_qty, na.rm = TRUE),
            total_sales_value = sum(transaction_qty * unit_price, na.rm = TRUE)) %>%
  arrange(desc(total_sales_value))

# Create a custom function for generating the bar plot
plot_sales_categories <- function(data, y_var, title, y_label, y_breaks_by, label_prefix = "") {
  ggplot(data, aes(x = reorder(product_category, -get(y_var)), y = get(y_var), fill = product_category)) +
    geom_bar(stat = "identity") +
    geom_label(aes(label = paste(label_prefix, round(get(y_var)), sep = "")),
               fill = "white", 
               size = 3, 
               fontface = "bold", 
               label.size = 0.3,
               vjust = -0.2,
               position = position_dodge(width = 0.9)) +
    scale_y_continuous(labels = comma, 
                       breaks = seq(0, max(data[[y_var]]) * 1.2, by = y_breaks_by),
                       limits = c(0, max(data[[y_var]]) * 1.2)) +
    labs(title = title, 
         x = "Product Category", 
         y = y_label) +
    theme_custom +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")
}

# Call the function for total sales quantity
plot_sales_categories(sales_by_category, 
              y_var = "total_sales_qty",
              title = "Total Sales Quantity by Product Category", 
              y_label = "Total Sales Quantity", 
              y_breaks_by = 10000, 
              label_prefix = "n = ")

# Call the function for total sales value
plot_sales_categories(sales_by_category, 
              y_var = "total_sales_value",
              title = "Total Sales Income by Product Category", 
              y_label = "Total Sales Income", 
              y_breaks_by = 25000, 
              label_prefix = "$")

```

``` {r}
# Summarize average sales per transaction by product category
avg_sales_per_transaction_category <- data_sales %>%
  group_by(product_category) %>%
  summarise(
    avg_sales_per_transaction = sum(transaction_qty * unit_price, na.rm = TRUE) / n()
  ) %>%
  arrange(desc(avg_sales_per_transaction))

# Plot the average sales per transaction by product category
ggplot(avg_sales_per_transaction_category, aes(x = reorder(product_category, avg_sales_per_transaction), y = avg_sales_per_transaction)) +
  geom_bar(stat = "identity", fill = "pink") +
  coord_flip() +
  labs(title = "Average Sales Per Transaction by Product Category", 
       x = "Product Category", 
       y = "Average Sales Per Transaction ($)") +
  geom_label(aes(label = scales::comma(round(avg_sales_per_transaction, 2))), 
             position = position_stack(vjust = 0.5), 
             size = 3, fill = "white") +
  theme_custom

```

Coffee beans and Branded products may be higher-priced items, leading to higher per-transaction sales even though they have fewer units sold.
Coffee and Tea are likely lower-priced items but are sold in much larger quantities, contributing more significantly to the overall revenue, even though their average sales per transaction are lower.

**Key Observations:**

1.  *Coffee* is the top-performing product category in both quantity and value. It has the highest number of units sold and also generates the most revenue.

2.  *Tea* is the second most popular category, both in terms of the number of units sold and total revenue, though its sales figures are significantly lower than Coffee.

3.  *Bakery* and *Drinking Chocolate* show moderate sales in both quantity and value, ranking third and fourth in both metrics.

There’s a clear concentration of sales in a few major categories (Coffee, Tea, and Bakery), with the other categories contributing much smaller amounts.

### 4.1.2.   Sales By Product Type

There are 29 distinct product types in the dataset. To streamline the analysis, I chose to focus on the product types that belong to the top-selling product categories. These top categories were identified as those contributing at least 10% of the total revenue. This threshold is flexible and can be adjusted as needed. 

``` {r product type}
# Calculate total sales income across all categories
total_sales_income <- sum(sales_by_category$total_sales_value, na.rm = TRUE)

# Filter product categories that contribute at least 10% of the total income
significant_categories <- sales_by_category %>%
  filter(total_sales_value >= 0.1 * total_sales_income) %>%
  pull(product_category)

# Filter the data for the most valuable product_category and aggregate sales by product_type
sales_by_type_significant <- data_sales %>%
  filter(product_category %in% significant_categories) %>%
  group_by(product_category, product_type) %>%
  summarise(
    total_sales_qty = sum(transaction_qty, na.rm = TRUE),
    total_sales_value = sum(transaction_qty * unit_price, na.rm = TRUE),
    count_transactions = n()  # Count of transactions for each product type
  ) %>%
  arrange(desc(total_sales_value))

# Custom function to create dodged bar plots for product types within categories
plot_sales_categories <- function(data, x_var, y_var, fill_var, title, y_label, y_breaks_by, label_prefix = "") {
  ggplot(data, aes(x = reorder(get(x_var), -get(y_var)), y = get(y_var), fill = get(fill_var))) +
    geom_bar(stat = "identity", position = position_dodge2(width = 0.9)) +
    geom_text(aes(label = paste(label_prefix, round(get(y_var)), sep = ""), 
                  y = get(y_var) / 2),
              size = 3, 
              fontface = "bold", 
              colour = "gray90",
              position = position_dodge2(width = 0.9), 
              vjust = 0.5) +  # Dodged labels to match the bars
    scale_y_continuous(labels = comma, 
                       breaks = seq(0, max(data[[y_var]]) * 1.2, by = y_breaks_by),
                       limits = c(0, max(data[[y_var]]) * 1.2)) +
    labs(title = title, 
         x = "Product Category", 
         y = y_label,
         fill = "Product Type") +
    coord_flip()+
    theme_custom +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "right")
}

# Call the function for total sales quantity (dodged by product type)
plot_sales_categories(sales_by_type_significant, 
                      x_var = "product_category", 
                      y_var = "total_sales_qty", 
                      fill_var = "product_type", 
                      title = "Total Sales Quantity by Product Category and Product Type", 
                      y_label = "Total Sales Quantity", 
                      y_breaks_by = 2500,
                      label_prefix = "n = ")

# Call the function for total sales value (dodged by product type)
plot_sales_categories(sales_by_type_significant, 
                      x_var = "product_category", 
                      y_var = "total_sales_value", 
                      fill_var = "product_type", 
                      title = "Total Sales Value by Product Category and Product Type", 
                      y_label = "Total Sales Value", 
                      y_breaks_by = 10000,
                      label_prefix = "$")

```

The analysis of sales by product type of the best selling categories shows:

The top three best-selling product types were *Brewed Chai Tea (sold 26,250 times), Gourmet Brewed Coffee (sold 25,973 times), and Barista Espresso (sold 24,943 times)*.
The top three product types generating the highest profit were *Barista Espresso (\$91,406), Brewed Chai Tea (\$77,082), and Hot Chocolate (\$72,416 from 17,457 sales)*.

It is important to note that the best-selling product type did not generate the highest profit, but still are in top three product types generating the highest profit. Let's ensure there is correlation between products sold and total sales value.

``` {r correlation between units sold and profit, width = 24, height = 6}
# Create scatter plot showing the relationship between units sold and total sales value (profit)
ggplot(sales_by_type_significant, aes(x = total_sales_qty, y = total_sales_value)) +
  geom_point(aes(color = product_type), size = 4, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "pink") +
  labs(title = "Correlation Between Units Sold and Total Profit by Product Type",
       x = "Units Sold",
       y = "Total Sales Value ($)",
       color = "Product Type") +
  theme_custom +
  theme(
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),
    legend.key.size = unit(0.1, "cm")
  )
```

The pink line represents the trend line (using linear regression), showing the general positive relationship between units sold and total sales value. This means that, as the number of units sold increases, total sales value generally increases as well. Since The correlation between sales and profit is positive, but the relationship isn't perfect—some products deviate from the trend.
No product types were discovered that had low sales but high profit, indicating that products that generate significant revenue generally also have high sales volumes.

**Ensuring that all stores are regularly stocked with these high-performing products is crucial, as they play a significant role in both sales volume and revenue.**

### 4.1.3.   Sales By Product ID

There are 80 unique product IDs in the dataset, but for this analysis, I concentrated on the Top 10 items that contribute the most significant portion of the total profit or were best-sellers. 

``` {r product id}
# Calculate the total profit for each product_id
sales_profit <- data_sales %>%
  group_by(product_id, product_detail) %>%
  summarise(
    total_sales_qty = sum(transaction_qty, na.rm = TRUE),
    total_sales_value = sum(transaction_qty * unit_price, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  arrange(desc(total_sales_value))  # Sort by total sales value in descending order

# Select the top 20 items contributing the most profit 
top_20_products_sales <- sales_profit %>%
  slice_max(order_by = total_sales_value, n = 10)

# Select the top 20 items by quantity sold
top_20_products_sales <- sales_profit %>%
  arrange(desc(total_sales_qty)) %>%
  slice(1:20)

ggplot(top_20_products_sales, aes(x = reorder(product_id, total_sales_value), y = total_sales_value, fill = product_detail)) +
  geom_bar(stat = "identity") +
  coord_flip() +  # Flip coordinates for a horizontal bar chart
  labs(title = "Top 20 Products by Total Profit",
       x = "Product ID",
       y = "Total Sales Value ($)",
       fill = "Product Detail") +
  theme_custom +
  theme(
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm")
  )

ggplot(top_20_products_sales, aes(x = reorder(product_id, total_sales_qty), y = total_sales_qty, fill = product_detail)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 20 Products by Sales Quantity",
       x = "Product ID",
       y = "Total Quantity",
       fill = "Product Detail") +
  theme_custom +
  theme(
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm")
  )
```
Since the previous analysis didn't provide sufficient insights, I decided to focus on identifying which drink sizes were the best sellers within the coffee, tea, and drinking chocolate categories. 

``` {r}
# Extract drink size from product_detail (assuming size is at the end)
sales_by_size <- data_sales %>%
  filter(product_category %in% c("Coffee", "Tea", "Drinking Chocolate")) %>%
  mutate(
    drink_size = case_when(
      grepl("Sm", product_detail) ~ "Small",
      grepl("Rg", product_detail) ~ "Regular",
      grepl("Lg", product_detail) ~ "Large",
      TRUE ~ "Unknown"
    )
  ) %>%
  group_by(product_category, drink_size) %>%
  summarise(total_sales_qty = sum(transaction_qty, na.rm = TRUE),
            total_sales_value = sum(transaction_qty * unit_price, na.rm = TRUE)) %>%
  arrange(desc(total_sales_qty))

# Plot the total sales by drink size for each category
ggplot(sales_by_size, aes(x = product_category, y = total_sales_qty, fill = drink_size)) +
  geom_bar(stat = "identity", position = "dodge2") +
  labs(title = "Best-Selling Drink Sizes by Category", 
       x = "Product Category", 
       y = "Total Sales Quantity",
       fill = "Drink Size") +
  theme_custom

ggplot(sales_by_size, aes(x = product_category, y = total_sales_value, fill = drink_size)) +
  geom_bar(stat = "identity", position = "dodge2") +
  labs(title = "Total Sales Value by Drink Size and Category", 
       x = "Drink Size", 
       y = "Total Sales Value ($)",
       fill = "Drink Size") +
  theme_custom

```

This visualization helps highlight which drink sizes are most popular in each product category, with Large and Regular sizes standing out across all categories.

These insights suggest that stocking and promoting Large and Regular drinks should be a priority for maximizing both sales volume and profit in these categories. Efforts should be focused on ensuring these sizes are always available, especially in the Tea and Drinking Chocolate categories.

## 4.2.   Sales by Store ID and Location
### 4.2.1.    Compare Different Stores by Sales Value and Quantity Sold

``` {r}
# Summarize sales by store ID
sales_by_store <- data_sales %>%
  group_by(store_location, store_id) %>%
  summarise(
    total_sales_qty = sum(transaction_qty, na.rm = TRUE),
    total_sales_value = sum(transaction_qty * unit_price, na.rm = TRUE)
  ) %>%
  arrange(desc(total_sales_value))
sales_by_store_long <- sales_by_store %>%
  pivot_longer(cols = c(total_sales_qty, total_sales_value), 
               names_to = "sales_metric", 
               values_to = "sales_value")

# Plot both total sales quantity and value in a faceted plot
ggplot(sales_by_store_long, aes(x = reorder(store_location, sales_value), y = sales_value, fill = store_location)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ sales_metric, scales = "free") +  # Create separate plots for each metric
  labs(title = "Total Sales Quantity and Value by Store Location", 
       x = "Store Location",
       y = "Sales Value or Quantity",
      fill = "Store Location") +
  theme_custom

```

### 4.2.2.    Sales by Product Category for Each Store Location

``` {r}
sales_by_category_stores <- data_sales %>%
  group_by(store_location, product_category) %>%
  summarise(
    total_sales_qty = sum(transaction_qty, na.rm = TRUE),
    total_sales_value = sum(transaction_qty * unit_price, na.rm = TRUE)
  ) %>%
  arrange(store_location, desc(total_sales_value))

ggplot(sales_by_category_stores, aes(x = reorder(product_category, total_sales_value), y = total_sales_value, fill = product_category)) +
  geom_bar(stat = "identity", position = "dodge2") +
  facet_wrap(~ store_location) +
  labs(title = "Sales by Product Category and Store Location (Profit)",
       x = "Product Category",
       y = "Total Sales Value ($)",
       fill = "Product Category") +
  theme_custom +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggplot(sales_by_category_stores, aes(x = reorder(product_category, total_sales_qty), y = total_sales_qty, fill = product_category)) +
  geom_bar(stat = "identity", position = "dodge2") +
  facet_wrap(~ store_location) +
  labs(title = "Sales by Product Category and Store Location (Sold Items)",
       x = "Product Category",
       y = "Total Sales Value ($)",
       fill = "Product Category") +
  theme_custom +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

### 4.2.3. Average Sales Per Transaction by Store

``` {r}
# Calculate average sales per transaction by store
avg_sales_per_transaction <- data_sales %>%
  group_by(store_location) %>%
  summarise(
    avg_sales_per_transaction = sum(transaction_qty * unit_price, na.rm = TRUE) / n()
  ) %>%
  arrange(desc(avg_sales_per_transaction))

# Plot average sales per transaction by store
ggplot(avg_sales_per_transaction, aes(x = reorder(store_location, avg_sales_per_transaction), y = avg_sales_per_transaction, fill = store_location)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Sales per Transaction by Store Location",
       x = "Store Location",
       y = "Average Sales per Transaction ($)",
       fill = "Store Location") +
  geom_label(aes(label = scales::comma(round(avg_sales_per_transaction, 2))), 
             position = position_stack(vjust = 0.5), 
             size = 5, fill = "white") +
  theme_custom
```

There is little variation in the quantity of products sold, total profit, and sales across different product categories among the three stores. Similarly, the average sales per transaction are quite consistent across the stores, indicating that customer spending patterns are generally similar.

## 4.5.   Sales by Timeline
